---
title: "Lab: Logistic Regression"
subtitle: "Princeton University"
author: "Suyog Chandramouli"
output: 
  tufte::tufte_html:
    css: 
    tufte_variant: "envisioned"
    highlight: github-dark
    fig_height: 10
    fig_width: 16
    toc: true
    toc_depth: 1
execute: 
  message: false
  warning: false
format: 
  html:
    code-fold: true
    code-overflow: wrap
engine: knitr
---

-   Assignment requirements:

    -   If you are using Github (recommended), make sure to commit and push your work to GitHub regularly, at least after each exercise. Write short and informative commit messages, and share the link to your assignment with me. If not, you can also send me the **rmd** & **rendered file** via Canvas.

    -   In this assignment, you will not need to code from scratch. Rather, you'll need to fill in code where needed. This assignment walks through a logistic regression implementation from EDA to model comparison — useful as a reference for your own future analyses.

    -   I want the assignments to begin reflecting a bit more of how you'd be doing things on your own, where you have some prior knowledge and you figure other things out (by referring to documentation, etc.). In addition to the rendered file, I also want you to submit **notes** of anything new that you learn while finishing the assignment, and any pain-points — we'll discuss more.

-   Note:

    -   If you are fitting a model, display the model output in a neatly formatted table. (The `gt`, `tidy`, and `kable` functions can help!). `modelsummary` also works well: <https://vincentarelbundock.github.io/modelsummary/articles/modelsummary.html>

    -   Make sure that your plots are clearly labeled — all axes, titles, etc.
    
    -   When computing confidence intervals for odds ratios, always exponentiate the CI endpoints from the log-odds scale. Do **not** add/subtract SE on the OR scale directly.

## Data: General Social Survey

The General Social Survey (GSS) has been used to measure trends in attitudes and behaviors in American society since 1972. In addition to collecting demographic information, the survey includes questions used to gauge attitudes about government spending priorities, confidence in institutions, lifestyle, and many other topics. A full description of the survey may be found here.

The data for this lab are from the 2016 General Social Survey. The original data set contains 2867 observations and 935 variables. We will use and abbreviated data set that includes the following variables:

natmass: Respondent’s answer to the following prompt:

“We are faced with many problems in this country, none of which can be solved easily or inexpensively. I’m going to name some of these problems, and for each one I’d like you to tell me whether you think we’re spending too much money on it, too little money, or about the right amount…are we spending too much, too little, or about the right amount on mass transportation?”

age: Age in years.

sex: Sex recorded as male or female

sei10: Socioeconomic index from 0 to 100

region: Region where interview took place

polviews: Respondent’s answer to the following prompt:

“We hear a lot of talk these days about liberals and conservatives. I’m going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal - point 1 - to extremely conservative - point 7. Where would you place yourself on this scale?”

The data are in gss2016.csv in the data folder.

## EDA

> Let’s begin by making a binary variable for respondents’ views on spending on mass transportation. Create a new variable that is equal to “1” if a respondent said spending on mass transportation is about right and “0” otherwise. Then plot the proportion of the response variable, using informative labels for each category.

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(modelsummary)
library(tidyr)
library(knitr)
library(easystats)
library(broom)
library(emmeans)
library(marginaleffects)
library(performance)
install.packages("arm")
library(arm)
```

```{r}
# load data
data <- read.csv("gss2016.csv")

```

Fill in the "\_\_\_\_" below to encode the binary variable

```{r}
data <- data %>%
   mutate(mass_trans_spend_right = ifelse(natmass == "About right", 1, 0))
```

```{r}
#Get proportions
mass_spend_summary <- data %>%
  count(mass_trans_spend_right) %>%
  mutate(proportion = n / sum(n))

#Change structure if needed
mass_spend_long <- mass_spend_summary %>%
  pivot_longer(cols = proportion, 
               names_to = "variable", 
               values_to = "proportion")

#Factorise for plot
mass_spend_long$mass_trans_spend_right <- as.factor(mass_spend_long$mass_trans_spend_right)

#Make plot
ggplot(mass_spend_long, aes(x = "", y = proportion, fill = factor(mass_trans_spend_right))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "", y = "Proportion", fill = "Mass Transit Spending About Right") +
  theme_minimal()
```

> Recode `polviews` so it is a factor with levels that are in an order that is consistent with question on the survey. Note how the categories are spelled in the data.

```{r}

data <- data %>%
  mutate(polviews = factor(polviews,
                           levels = c("Extremely liberal",
                                      "Liberal",
                                      "Slightly liberal",
                                      "Moderate",
                                      "Slightly conservative",
                                      "Conservative",
                                      "Extremely conservative"),
                           ordered = TRUE))
```

> Make a plot of the distribution of `polviews`

```{r}
#Get proportions, format, and produce a plot like you did previously for mass_trans_spend_right
polviews_summary <- data %>%
  count(polviews) %>%
  mutate(proportion = n / sum(n))

polviews_long <- polviews_summary %>%
  pivot_longer(cols = proportion,
               names_to = "variable",
               values_to = "proportion")

polviews_long$polviews <- factor(polviews_long$polviews,
                                  levels = c("Extremely liberal",
                                             "Liberal",
                                             "Slightly liberal",
                                             "Moderate",
                                             "Slightly conservative",
                                             "Conservative",
                                             "Extremely conservative"),
                                  ordered = TRUE)

ggplot(polviews_long, aes(x = "", y = proportion, fill = polviews)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "", y = "Proportion", fill = "Political Views") +
  theme_minimal()
  
```

> Which political view occurs most frequently in this data set?
>
> Moderate views

> Make a plot displaying the relationship between satisfaction with mass transportation spending and political views. Use the plot to describe the relationship the two variables.

```{r}

plot_data <- data %>%
  filter(!is.na(polviews), !is.na(mass_trans_spend_right)) %>%
  group_by(polviews) %>%
  summarise(
    prop_right = mean(mass_trans_spend_right),
    n = n()
  )

ggplot(plot_data, aes(x = polviews, y = prop_right, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "steelblue") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = "Political Views",
    y = "Proportion saying spending is 'about right'",
    title = "Satisfaction with Mass Transit Spending by Political Views"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The more conservative one's political views are the more they think the amount of spending on mass transportation is correct.

-   We’d like to use age as a quantitative variable in your model; however, it is currently a character data type because some observations are coded as "89 or older".

> Recode age so that is a numeric variable. Note: Before making the variable numeric, you will need to replace the values "89 or older" with a single value.

```{r}

data <- data %>%
  mutate(age = if_else(age == "89 or older", "89", age),
         age = as.numeric(age))
```

> Plot the frequency distribution of age.

```{r}
ggplot(data, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(x = "Age (years)", y = "Frequency",
       title = "Frequency Distribution of Age") +
  theme_minimal()
```

## Logistic regression

> Let’s start by fitting a logistic regression model with just the intercept

```{r}
intercept_only_model <- glm(
  mass_trans_spend_right ~ 1,
  data = data,
  family = binomial(link = "logit")
)

intercept_only_model %>% 
  tidy() %>%
  kable()

```

> Interpret the intercept in the context of the data. You can do this by converting the $\beta_0$ parameter out of the log-odds metric to the probability metric. Make sure to include the 95% confidence intervals. Then interpret the results in a sentence or two--what is the basic thing this probability tells us about?

```{r}


b0 <- coef(intercept_only_model)[1]  # get coef
b0_transformed <- plogis(b0)  # convert log-odds to probability

# CIs on log-odds scale, then convert
ci_lower_logodds <- b0 - 1.96 * 0.0393685
ci_upper_logodds <- b0 + 1.96 * 0.0393685

p_lower <- plogis(ci_lower_logodds)
p_upper <- plogis(ci_upper_logodds)


b0_transformed
p_lower
p_upper

```

Interpretation: The estimate is very close to 50%, which means satisfaction with mass transit spending is roughly split in the population. The CI is small so we are pretty confident this is close to the truth for the population. 



> Now let's fit a model using the demographic factors - `age`,`sex`, `sei10` - to predict the odds a person is satisfied with spending on mass transportation. Make any necessary adjustments to the variables so the intercept will have a meaningful interpretation. *Neatly display the model coefficients (do not display the summary output)*

*Hint: For the intercept to be meaningful, it should represent the log-odds for a realistic person. What does age = 0 and sei10 = 0 mean? Consider centering continuous predictors.*

```{r}
# Make sure sex is a factor
data <- data %>%
  mutate(sex = as.factor(sex))

# Center continuous predictors so the intercept is meaningful
data <- data %>%
  mutate(age_c = age - mean(age, na.rm = TRUE),
         sei10_c = sei10 - mean(sei10, na.rm = TRUE))

# Fit with glm()
m1 <- glm(
  mass_trans_spend_right ~ age_c + sex + sei10_c,
  data = data,
  family = binomial(link = "logit")
)

# Produce tidy output of model coefficients
m1 %>%
  tidy() %>%
  kable()
```

> Consider the relationship between sex and one’s opinion about spending on mass transportation. Interpret the coefficient of sex in terms of the log-odds and OR of being satisfied with spending on mass transportation. What are the predicted probabilities for males and females on support for spending on mass transportation? Please include the 95% CIs around each estimate.

Males have log-odds 0.256 lower than females, holding age and SEI constant. exp(-0.256) ≈ 0.774, so the odds of being satisfied for males are about 77.4% of the odds for females — roughly a 23% reduction in odds.

```{r}
m1 %>% 
  tidy() %>%
  kable()

m1 %>% 
  tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  kable()

bsex <- coef(m1)["sexFemale"]

ci_lower_lo = bsex - 1.96 * 0.0798020
ci_upper_lo = bsex + 1.96 * 0.0798020



# exponentiate the log-odds CIs
ci_lower_or = exp(ci_lower_lo)
ci_upper_or = exp(ci_upper_lo)

# Get predicted probabilities with emmeans
emm_sex <- emmeans(m1, "sex", type = "response")
emm_sex

```

If you did this right, you'll find that being female (as compared to male) is associated with an increase in the log-odds of being satisfied with spending on mass transportation by 0.2557439 units (95% CI \[0.09, 0.41\]), holding all other variables constant. This equates to the odds of thinking the spending amount is right in females being 1.29 times the odds of thinking this in men (95% CI \[1.13, 1.44\]).

The predicted probability for females to be satisfied with spending on mass transportation is 55.9% (95% CI \[53.3%, 58.5%\]) and that of males is 49.5% (95% CI \[46.7%, 52.4%\]).\
\
Verify this. 
Yes got the same thing above!

> Next, consider the relationship between age and one’s opinion about spending on mass transportation. Interpret the coefficient of age in terms of the log-odds and OR of being satisfied with spending on mass transportation. Please include the 95% CIs around each estimate.

```{r}

bage <- coef(m1)["age_c"]  
se_age <- summary(m1)$coefficients["age_c", "Std. Error"]

# Log-odds CIs
ci_lower_lo <- bage - 1.96 * se_age
ci_upper_lo <- bage + 1.96 * se_age

# OR and OR CIs (exponentiate the log-odds CIs)
or_age <- exp(bage)
ci_lower_or <- exp(ci_lower_lo)
ci_upper_or <- exp(ci_upper_lo)

```

A one unit increase in age is associated with a decrease in the log-odds of being satisfied with spending on mass transportation by 0.0062 (95% CI [-0.0106, -0.0017]), holding all other variables constant. The odds ratio is 0.9939 (95% CI [0.9894, 0.9983]), which confirms the negative relationship implied by the log-odds coefficient. Specifically, for each additional unit of age, the odds of being satisfied with mass transportation spending decrease by a factor of about 0.994, or approximately 0.6% per unit increase in age, holding other factors constant.

> Consider the relationship between SES and one’s opinion about spending on mass transportation. Interpret the coefficient of SES in terms of the log-odds and OR of being satisfied with spending on mass transportation. Please include the 95% CIs around each estimate.

```{r}
bses <- coef(m1)["sei10_c"]   # use "sei10_c" if you centered it
se_ses <- summary(m1)$coefficients["sei10_c", "Std. Error"]

# Log-odds CIs
ci_lower_lo <- bses - 1.96 * se_ses
ci_upper_lo <- bses + 1.96 * se_ses

# OR and OR CIs
or_ses <- exp(bses)
ci_lower_or <- exp(ci_lower_lo)
ci_upper_or <- exp(ci_upper_lo)
```


A one unit increase in SES index is associated with a decrease in the log-odds of being satisfied with spending on mass transportation by 0.0062 units (95% CI [-0.0095, -0.0030]), holding all other variables constant. The odds ratio is 0.9938 (95% CI [0.9906, 0.9970]), which confirms the negative relationship implied by the log-odds coefficient. Specifically, for each additional unit of SES index, the odds of being satisfied with mass transportation spending change by a factor of about 0.994, or approximately 0.6% per unit increase in SES index, holding other factors constant.

## Marginal effects

-   Let's examine the results on the probability scale.

> Calculate the marginal effects of sex, age, and SES on mass transportation spending. You can use the `margins` package function `margins` discussed in your textbook or you can use the `marginaleffects` package `avg_slope` `avg_comparisons` discussed in lecture. Interpret each estimate.

```{r}
avg_comparisons(m1, comparison = "difference") %>% 
  kable()

```

- The marginal effect of age is -0.0015 (95% CI [-0.0026, -0.0004]). So, for each additional unit increase of age, the probability of being satisfied with mass transportation spending decreases by approximately 0.15 percentage points, holding other factors constant (p = .007).

- The marginal effect of SES is -0.0015 (95% CI [-0.0023, -0.0007]). For each one-unit increase in the socioeconomic index, the probability of being satisfied with mass transportation spending decreases by approximately 0.15 percentage points, holding other variables constant.

- The marginal effect for being female compared to male is 0.0631 (95% CI [0.0246, 0.1016]). This indicates that females are, on average, about 6.3 percentage points more likely than males to be satisfied with mass transportation spending, holding other factors constant.

## Model comparison

-   Now let’s see whether a person’s political views has a significant impact on their odds of being satisfied with spending on mass transportation, after accounting for the demographic factors.

> Conduct a drop-in-deviance/likelihood ratio test to determine if polviews is a significant predictor of attitude towards spending on mass transportation. Name these two models `fit2` and `fit3`, respectively. Compare the two models.

```{r}

data_complete <- data %>%
  filter(!is.na(mass_trans_spend_right),
         !is.na(age_c), !is.na(sex), !is.na(sei10_c),
         !is.na(polviews))

fit2 <- glm(
  mass_trans_spend_right ~ age_c + sex + sei10_c,
  data = data_complete,
  family = binomial(link = "logit")
)

fit3 <- glm(
  mass_trans_spend_right ~ age_c + sex + sei10_c + polviews,
  data = data_complete,
  family = binomial(link = "logit")
)

test_likelihoodratio(fit2, fit3) %>% kable()

```

> Is the model with polviews better than the model without?

-   Yes.

## Visualization

-   Let’s plot the results

-   We next use the model to produce visualizations:

    -   Given the code below, interpret what is being plotted:

        -   pol_plot : \_\_\_\_

        -   sex_plot : \_\_\_\_

        -   ses_plot: \_\_\_

    ::: callout-tip
    -   adjust the various settings in your plot to make it look professional.

    -   You can use `ggeffects` to get the predicted probabilities for these models.
    :::

```{r}
#| fig-width: 12
#| fig-height: 10
library(ggeffects)


colors <- c("Extremely liberal" = "black",
            "Liberal" = "#0e2f44",  # Dark blue
            "Slightly liberal" = "#1d5a6c",  # Less dark blue
            "Moderate" = "#358ca3",  # Medium blue
            "Slghtly conservative" = "#71b9d1",  # Light blue
            "Conservative" = "#a6dcef",  # Lighter blue
            "Extrmly conservative" = "#d0f0fd")  # Very light blue

pp_pol <- ggemmeans(fit3, terms = c("polviews"))

# Adjusted plot with gradient colors
pol_plot <- ggplot(pp_pol, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  scale_color_manual(values = colors) +
  labs(title = "Effect of Political Views on Satisfaction with Mass Transportation",
       x = "Political Views", y = "Predicted Probability",
       color = "Political Views") +
  theme_minimal()

pol_plot

pp_sex <- ggemmeans(fit3, terms = c("sex"))

sex_plot <- ggplot(pp_sex, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(title = "Effect of Sex on Satisfaction with Mass Transportation",
       x = "Sex", y = "Predicted Probability",
       color = "Sex") +
  theme_minimal()

pp_sex

pp_ses <- ggemmeans(fit3, terms = "sei10_c")


ses_plot <-  ggplot(pp_ses, aes(x = x, y = predicted)) +
  geom_line(color = "#2c7fb8", size = 1) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#2c7fb8", alpha = 0.2) +  # Add a confidence interval band
  labs(title = "Effect of SES on Satisfaction with Mass Transportation",
       x = "Socioeconomic Status", y = "Predicted Probability") +
  theme_minimal() +
  theme(legend.position = "none")  
ses_plot

```

## Model Assumptions

-   Is the logistic model a good choice for this data?

```{r}
binned_residuals(fit2)
```

::: callout-note
Answer: The model is interpretable and the fact that 91% of residuals are in the error bounds means is pretty good!
:::

## Model fit

-   Calculate the $R^2$ for this model

```{r}

r2_mcfadden(fit2)

```

> R2 interpretation: This is ok but not great by any means. For the McFadden R2 0.12 is ok but ideally it would be over .2
>
> Next, Take a look at the binned residual plots for each continuous predictor variable and look at linearity. Is there a predictor that sticks out? What can we do to improve model fit in this case?

```{r}
binned_residuals(fit2, term="sei10_c")
binned_residuals(fit2, term="age_c")

binned_residuals(fit2, term="sei10_c") %>% plot(show_dots=TRUE)
binned_residuals(fit2, term="age_c") %>% plot(show_dots=TRUE)

```

::: callout-note

SEI seems to be having issues and is the predictor that sticks out. 
:::

## Improving Model Fit

> Based on your binned residual plots above, one predictor may show signs of nonlinearity. Try adding a squared term for that predictor, refit the model, and compare using a likelihood ratio test. Do the binned residuals improve?

```{r}
# 
fit2_improved <- glm(
  mass_trans_spend_right ~ age_c + sex + poly(sei10_c, 2) + polviews,
  data = data_complete,
  family = binomial(link = "logit")
)

#the other didnt work
anova(fit2, fit2_improved, test = "LRT")

binned_residuals(fit2_improved, term = "sei10_c") %>% plot(show_dots = TRUE)

```

::: callout-note
Did adding the squared term improve the model? According to the anova it is better, but the plot still shows lots of points outside the bounds. 
:::

## Testing Polviews

```{r}
emmeans(fit3, "polviews") %>% pairs() %>% as.data.frame() %>% filter(p.value < .05)

emmeans(fit3, "polviews", type="response") %>% pairs() %>% as.data.frame() %>% filter(p.value < .05)
```

> Using the pairwise comparisons above, fill in the blanks. Be specific: are these odds ratios or probability differences? The `type="response"` output gives odds ratios for logistic models.

-   Conservatives have ~2.71 (1/0.3684) times the odds compared to extremely liberal, and ~2.17 (1/0.4604) times the odds compared to liberal respondents.

-   Extreme liberals have ~0.37, ~0.40, and ~ [slightly conservative not in output] times the odds compared to conservatives, moderates, and slightly conservative respondents respectively (extreme liberals have lower odds).

-   Extremely conservative respondents — not in your output (table only shows Extremely liberal, Liberal, Moderate, Conservative); you'd need to run pairwise comparisons including that level to fill this in.

-   Liberals have ~0.50 times the odds compared to moderates, and ~ [slightly conservative not in output] — again, slightly conservative isn't in this table.


## Conclusion

The full model, including age, sex, SEI, and political views, provides a significantly better fit than the null model, based on the deviance table above and the three figures. Each predictor contributes significantly and each reduce deviance meaningfully, and political views adds the largest improvement meaning that ideology is the strongest predictor of satisfaction with mass transit spending. Based on the figs above, some nonlinearity in SEI remains




```{r model-summary, results='asis', echo=FALSE , fig.cap="Table 1"}

fit_anova = anova(fit3, test="Chisq")

fit_anova %>%
  kable()


```

```{r plot-sexviews, fig.cap="Figure 1: Effect of Sex on Satisfaction with Mass Transportation", echo=FALSE}
sex_plot
```

```{r plot-sesviews, fig.cap="Figure 2: Effect of SES  on Satisfaction with Mass Transportation", echo=FALSE}
ses_plot
```

```{r plot-polviews, fig.cap="Figure 3: Effect of Political Views on Satisfaction with Mass Transportation", echo=FALSE}
pol_plot
```
