---
    title: "Lab 4 - Multinomial Regression"
    author:
      - name: Suyog Chandramouli
    date: last-modified
    format:
      html:
        self-contained: true
        anchor-sections: true
        code-tools: true
        code-fold: true
        fig-width: 8
        fig-height: 4
        code-block-bg: "#f1f3f5"
        code-block-border-left: "#31BAE9"
        mainfont: Source Sans Pro
        theme:
          - journal
          - custom.scss
        toc: true
        toc-depth: 3
        toc-location: left
        captions: true
        cap-location: margin
        table-captions: true
        tbl-cap-location: margin
        reference-location: margin
      pdf:
        pdf-engine: lualatex
        toc: false
        number-sections: true
        number-depth: 2
        top-level-division: section
        reference-location: document
        listings: false
        header-includes:
          \usepackage{marginnote, here, relsize, needspace, setspace}
          \def\it{\emph}

    comments:
      hypothesis: false

    execute:
      warning: false
      message: false
---

**Lab Goal:** Predict voting frequency using demographic variables\
**Data source:** FiveThirtyEight "Why Many Americans Don't Vote" survey\
**Method:** Multinomial logistic regression

## Instructions

-   If you are fitting a model, display the model output in a neatly formatted table. (The `tidy()` and `kable()` functions can help!)
-   If you are creating a plot, use clear labels for all axes, titles, etc.
-   Commit and push your work to GitHub regularly, at least after each exercise. Write short and informative commit messages.
-   When you're done, we should be able to knit the final version of the QMD in your GitHub as an HTML.

::: callout-tip
## Workflow tip

Work through the lab by running chunks **one at a time** in RStudio (Cmd/Ctrl + Enter), not by rendering the whole document. Render only at the end, once all your code is working.

This is how most R users actually work -- you'd write a line, run it, check the output, and iterate. Rendering is for producing the final document. When you render, Quarto starts a fresh R session and runs every chunk top-to-bottom, so if anything is incomplete, everything downstream breaks. When you work interactively using chunks, your environment builds up as you go, so you always have your objects available to inspect and troubleshoot.

The **"Run All Chunks Above"** button (downward arrow with a bar, next to the green play button in each chunk) is another useful tool. It re-runs everything from the top of the document up to your current chunk, so all your objects are up to date before you start working on the next question.
:::

## Setup

```{r}
#| code-fold: false
library(nnet)
library(car)
library(tidyverse)
library(emmeans)
library(ggeffects)
library(knitr)
library(patchwork)
library(broom)
library(parameters)
library(easystats)
library(viridis)
```

------------------------------------------------------------------------

# Part 1: Data Preparation

## The Data

The data for this assignment comes from an online Ipsos survey that was conducted for the FiveThirtyEight article ["Why Many Americans Don't Vote"](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/). If you'd like, you can read more about the survey design and respondents in the README of the [GitHub repo](https://github.com/fivethirtyeight/data/tree/master/non-voters) for the data, but all the necessary information is explained below.

Respondents were asked a variety of questions about their political beliefs, thoughts on multiple issues, and voting behavior. We will focus on using the demographic variables and someone's party identification to understand whether a person is a probable voter.

The variables we'll focus on are (definitions from the codebook in data set GitHub repo):

-   `ppage`: Age of respondent

-   `educ`: Highest educational attainment category

-   `race`: Race of respondent, census categories. Note: all categories except Hispanic were non-Hispanic.

-   `gender`: Gender of respondent

-   `income_cat`: Household income category of respondent

-   `Q30`: Response to the question "Generally speaking, do you think of yourself as a..."

    -   1: Republican
    -   2: Democrat
    -   3: Independent
    -   4: Another party, please specify
    -   5: No preference
    -   -1: No response

-   `voter_category`: past voting behavior:

    -   **always**: respondent voted in all or all-but-one of the elections they were eligible in
    -   **sporadic**: respondent voted in at least two, but fewer than all-but-one of the elections they were eligible in
    -   **rarely/never**: respondent voted in 0 or 1 of the elections they were eligible in

```{r}
#| code-fold: false
voter_data <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv")
```

The variable `Q30` contains the respondent's political party identification. The following code simplifies `Q30` into four categories:

```{r}
#| code-fold: false
voter_data <- voter_data %>%
  mutate(pol_ident = case_when(
    Q30 == 1 ~ "Rep",
    Q30 == 2 ~ "Dem",
    Q30 == 3 ~ "Indep",
    TRUE ~ "Other"
  ))
```

### Question 1.1 {.unnumbered}

The variable `voter_category` identifies the respondent's past voter behavior. Relevel the variable to make **rarely/never** the baseline level, followed by sporadic, then always.

::: {.callout-tip collapse="true" title="Hint"}
Use `factor()` with the `levels` argument to set the ordering. The first level becomes the baseline in `multinom()`.
:::

```{r}
library(dplyr)

voter_data <- voter_data |> mutate(voter_category = factor(voter_category, levels = c("rarely/never", "sporadic", "always")))
```

### Question 1.2 {.unnumbered}

Mean-center the age variable (`ppage`) so that the intercept reflects the log-odds for an average-aged person rather than a 0-year-old.

```{r}
voter_data <- voter_data |> mutate(ppage_c = ppage - mean(ppage, na.rm = TRUE))
colnames(voter_data)

```

------------------------------------------------------------------------

# Part 2: Exploratory Data Analysis

### Question 2.1 {.unnumbered}

In the [FiveThirtyEight article](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/), the authors include visualizations of the relationship between voter category and demographic variables such as race, age, education, etc.

Select **two** demographic variables. For each variable, create a visualization showing its relationship with `voter_category`. Combine them into a single figure using `patchwork`. Briefly describe (1--2 sentences each) what the plots reveal.

For inspiration on recreating the FiveThirtyEight style: <https://www.mikelee.co/posts/2020-02-08-recreate-fivethirtyeight-chicklet-stacked-bar-chart-in-ggplot2>

::: {.callout-tip collapse="true" title="Hint: Quick proportional bar charts"}
`geom_bar(position = "fill")` gives proportions instead of counts. Map `x` to your demographic variable and `fill` to `voter_category`.
:::

```{r}

install.packages("remotes")   # if you don’t already have it
remotes::install_github("hrbrmstr/ggchicklet")

# load libraries
library(ggplot2)
library(ggchicklet)
library(ggtext)
library(dplyr)
library(forcats)
library(grid)

```



```{r}

voter_colors <- c(
  "always"       = "#72D4DB",
  "sporadic"     = "#B5BEC9",
  "rarely/never" = "#e84c4c"
)


race_dat <- voter_data %>%
  count(race, voter_category) %>%
  group_by(race) %>%
  mutate(Percent = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    race = fct_relevel(race, rev(unique(race))),
    voter_category = fct_relevel(
      voter_category,
      c("rarely/never", "sporadic", "always")
    )
  )

race_dat$Label <- paste0(sprintf("%.1f", race_dat$Percent * 100), "%")

gg_race <- ggplot(race_dat, aes(race, Percent, fill = voter_category, label = Label)) +
  geom_chicklet(
    width = 1,
    radius = unit(6, "pt"),
    position = position_stack(reverse = FALSE)
  ) +
  geom_text(
    size = 4,
    fontface = "bold",
    position = position_stack(vjust = 0.5)
  ) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_markdown(hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(
      hjust = 0,
      size = 11,
      margin = unit(c(0, 0, 0, 0), "cm")
    ),
    legend.text = element_markdown(size = 11),
    axis.text = element_text(face = "bold", size = 11),
    axis.text.x = element_blank(),
    axis.title.y = element_markdown(
      hjust = 0,
      size = 16,
      margin = unit(c(0, 0, 0, 0), "cm")
    ),
    panel.grid = element_blank(),
    axis.title.x = element_markdown(
      halign = 0,
      margin = margin(2, 0, 15, 0),
      fill = "transparent"
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = voter_colors,
    labels = c(
      "rarely/never" = "<strong>RARELY / NEVER</strong>",
      "sporadic"     = "<strong>SPORADIC</strong>",
      "always"       = "<strong>ALWAYS</strong>"
    ),
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    y = " ",
    x = NULL,
    fill = NULL,
    title = "<b>Voter participation by race</b>",

  )

gg_race

```

This shows the voter breakdown by race and we can see that White and Black Americans vote at higher rates than other groups. 


```{r}
educ_dat <- voter_data %>%
  count(educ, voter_category) %>%
  group_by(educ) %>%
  mutate(Percent = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    educ = fct_relevel(educ, rev(unique(educ))),
    voter_category = fct_relevel(
      voter_category,
      c("rarely/never", "sporadic", "always")
    )
  )

educ_dat$Label <- paste0(sprintf("%.1f", educ_dat$Percent * 100), "%")

gg_educ <- ggplot(educ_dat, aes(educ, Percent, fill = voter_category, label = Label)) +
  geom_chicklet(
    width = 1,
    radius = unit(6, "pt"),
    position = position_stack(reverse = FALSE)
  ) +
  geom_text(
    size = 4,
    fontface = "bold",
    position = position_stack(vjust = 0.5)
  ) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_markdown(hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(
      hjust = 0,
      size = 11,
      margin = unit(c(0, 0, 0, 0), "cm")
    ),
    legend.text = element_markdown(size = 11),
    axis.text = element_text(face = "bold", size = 11),
    axis.text.x = element_blank(),
    axis.title.y = element_markdown(
      hjust = 0,
      size = 16,
      margin = unit(c(0, 0, 0, 0), "cm")
    ),
    panel.grid = element_blank(),
    axis.title.x = element_markdown(
      halign = 0,
      margin = margin(2, 0, 15, 0),
      fill = "transparent"
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = voter_colors,
    labels = c(
      "rarely/never" = "<strong>RARELY / NEVER</strong>",
      "sporadic"     = "<strong>SPORADIC</strong>",
      "always"       = "<strong>ALWAYS</strong>"
    ),
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(
    y = " ",
    x = NULL,
    fill = NULL,
    title = "<b>Voter participation by education</b>",
  )


gg_educ
```
This shows the voter participant by aducation and we can see that more education leads to higher voter rates.


```{r}


alignTitles <- function(ggplot, title = NULL, subtitle = NULL, caption = NULL) {
  g <- ggplotGrob(ggplot)

  if (!is.null(title)) {
    g$layout[which(g$layout$name == "title"), ]$l <- title
  }
  if (!is.null(subtitle)) {
    g$layout[which(g$layout$name == "subtitle"), ]$l <- subtitle
  }
  if (!is.null(caption)) {
    g$layout[which(g$layout$name == "caption"), ]$l <- caption
  }
  g
}


grid.newpage()
grid.draw(alignTitles(gg_race, caption = 2))

grid.newpage()
grid.draw(alignTitles(gg_educ, caption = 2))
```

> Enter your interpretation here: It seems like high school or less votes less and also races other than Black and White Americans.

------------------------------------------------------------------------

# Part 3: Model Building

### Question 3.1 {.unnumbered}

Fit a multinomial logistic regression model using mean-centered age, race, gender, income, and education to predict `voter_category`. Save this as `model_demo`.

You don't need to display the output here — we'll examine it in Part 4. If you want to peek at it, feel free to run `summary(model_demo)` in your console. There's no real harm in displaying it via the code chunk, except that it'll make your rendered page quite messy. `kable()` is better for clean tables.

::: {.callout-tip collapse="true" title="Hint"}
`multinom(voter_category ~ age_centered + race + gender + income_cat + educ, data = voter_data, trace = FALSE)` — the `trace = FALSE` suppresses the iteration output.

To keep the rendered document clean, add `#| results: hide` to the chunk header — this runs the code but hides the output. You can always check the output interactively by running the chunk in the console.
:::

```{r}
#| results: hide
voter_data <- voter_data |>  mutate(age_centered = ppage - mean(ppage, na.rm = TRUE))

model1 <- multinom(voter_category ~ age_centered + race + gender + income_cat + educ, data = voter_data, trace = FALSE)

```

### Question 3.2 {.unnumbered}

Should party identification (`pol_ident`) be added to the model? Fit a second model that adds `pol_ident` to the predictors. Use `anova()` to compare the two models. Report the test statistic and your conclusion.

```{r}
model2 <- multinom(voter_category ~ age_centered + race + gender + income_cat + educ + pol_ident, data = voter_data, trace = FALSE)

anova(model1, model2)

```

> Enter your answer: Party identification should be used yes because the Pr(Chi) is 0 which means the result is significant. 

**Use the model you select for the remainder of the assignment.** For clarity, call your chosen model `final_model`.

```{r}
final_model <- multinom(voter_category ~ age_centered + race + gender + income_cat + educ + pol_ident, data = voter_data, trace = FALSE)

```

------------------------------------------------------------------------

# Part 4: Interpreting the Model

### Question 4.1: Overall significance {.unnumbered}

Use a likelihood ratio test (`car::Anova()` with `type = "III"`) to report the overall significance of each predictor in `final_model`.

```{r}
Anova(final_model, type = "III")
```

### Question 4.2: Raw coefficients {.unnumbered}

Display the model output using `tidy(final_model) %>% kable(digits = 3)`.

Pick **one** comparison (e.g., "sporadic vs. rarely/never") and do the following:

a)  Write out the model equation for that comparison.
b)  Interpret the coefficient for mean-centered age: what happens to the **odds** of being in that category (vs. the baseline) for each additional year of age? Report the odds ratio `exp(β)`.

::: {.callout-tip collapse="true" title="Hint: Reading the tidy() output"}
When you run `tidy(final_model)`, you'll see a table with columns like `y.level`, `term`, `estimate`, etc. The table has **two blocks of rows** — one block where `y.level` is `sporadic`, and another where `y.level` is `always`. Each block is a separate logistic regression equation:

-   The `sporadic` rows are the equation for: log-odds of sporadic **vs. rarely/never**
-   The `always` rows are the equation for: log-odds of always **vs. rarely/never**

Pick one of those two blocks. The `(Intercept)` row is $\beta_0$ and the `age_centered` row is $\beta_1$ for that equation.
:::

```{r}
tidy(final_model) %>% kable(digits = 3)

```

> Enter your equation and interpretation here: 

a) log( P(always) / P(rarely/never) ) = B0 (intercept) + Bn(variable_name)...

b)  The coefficient for mean-centered age: 0.058 which is in log odds. This means that for every additional year of age, the odds of you being in always vs never increases by (odds = 1.06) 6%.  Report the odds ratio `exp(β)` = 1 : 1.06

### Question 4.3: Changing the baseline {.unnumbered}

Refit the model with a **different** baseline category for `voter_category` (e.g., "always" instead of "rarely/never"). Then predict the probabilities for a specific observation using both models.

Do the coefficients change? Do the predicted probabilities change? Explain in 1--2 sentences why this happens.

::: {.callout-tip collapse="true" title="Hint"}
Use `relevel(voter_category, ref = "always")` before refitting. Use `predict(model, newdata = ..., type = "probs")` to get predicted probabilities. You can use `head(voter_data, 1)` as your newdata for a quick check.
:::

```{r}
voter_data <- voter_data |> mutate(voter_category = factor(voter_category, levels = c("always", "rarely/never", "sporadic")))

final_model <- multinom(voter_category ~ age_centered + race + gender + income_cat + educ + pol_ident, data = voter_data, trace = FALSE)
tidy(final_model) %>% kable(digits = 3)

```

> Enter your answer: With a new baseline the coefficients will change because they are with reference to a different baseline, however this does not change the predicted probabilities from going from each category to another, the model itself doesn't change. 

------------------------------------------------------------------------

# Part 5: Marginal Effects & Visualization

### Question 5.1: Predicted probabilities by age {.unnumbered}

Plot the predicted probabilities of voter category as a function of age. The code below will work once you replace `final_model` with the name of your model:

::: callout-note
This chunk is set to `eval: false` so the file renders before you've built your model. Once `final_model` exists in your environment, delete the `#| eval: false` line so the chunk runs when you render.
:::

```{r}

ggemmeans(final_model, terms = c("age_centered")) %>%
  ggplot(., aes(x = x, y = predicted, fill = response.level)) +
  geom_area() +
  geom_rug(sides = "b", position = "jitter", alpha = .2) +
  labs(x = "\nAge", y = "Predicted Probability\n",
       title = "Predicted Probabilities of Voting Frequency by Age") +
  scale_fill_manual(
    name = NULL,
    values = c("always" = "#F6B533", "sporadic" = "#D07EA2", "rarely/never" = "#9854F7"),
    labels = c("RARELY OR NEVER VOTE", "SOMETIMES VOTE", "ALMOST ALWAYS VOTE"),
    breaks = c("rarely/never", "sporadic", "always")
  ) +
  theme_minimal()
```

Describe the pattern you see in 1--2 sentences.

> Enter your interpretation here: As age increases the rarely or never category becomes incredibly small, whereas in younger voters it is more than 50% of the survey respondents. 


### Question 5.2: Predicted probabilities by education {.unnumbered}

Create a similar plot showing predicted probabilities of voter category as a function of **education level**.

::: {.callout-tip collapse="true" title="Hint"}
Replace `"ppage [all]"` with `"educ"` in the `ggemmeans()` call. Since education is categorical, you may want to use `geom_col(position = "dodge")` or `geom_col(position = "fill")` instead of `geom_area()`.
:::

```{r}

ggemmeans(final_model, terms = c("educ")) %>%
  ggplot(., aes(x = x, y = predicted, fill = response.level)) +
  geom_col(position = "dodge") +
  geom_rug(sides = "b", position = "jitter", alpha = .2) +
  labs(x = "\nAge", y = "Predicted Probability\n",
       title = "Predicted Probabilities of Voting Frequency by Age") +
  scale_fill_manual(
    name = NULL,
    values = c("always" = "#F6B533", "sporadic" = "#D07EA2", "rarely/never" = "#9854F7"),
    labels = c("RARELY OR NEVER VOTE", "SOMETIMES VOTE", "ALMOST ALWAYS VOTE"),
    breaks = c("rarely/never", "sporadic", "always")
  ) +
  theme_minimal()
```

> Enter your interpretation here: The almost alwasy vote is a lot higher for people with college education and the rarely votes goes down with more education. 

------------------------------------------------------------------------

# Part 6: Pairwise Comparisons

### Question 6.1: Political group differences {.unnumbered}

Run the following code (replacing `final_model` with your model name) and interpret the output. The code computes estimated marginal means for each political group within each voter category, then tests pairwise differences on the log-odds scale.

::: callout-note
Same as Q5.1 — this chunk is set to `eval: false`. Delete that line once `final_model` exists.
:::

```{r}

# Estimated marginal means: political group by voter category
multi_an_pol <- emmeans(final_model, ~ pol_ident | voter_category)

# Contrasts on the log scale (log-odds differences between voter categories, within each political group)
coefs_pol <- contrast(regrid(multi_an_pol, "log"), "trt.vs.ctrl1", by = "pol_ident")

# Display results, organized by contrast
update(coefs_pol, by = "contrast") %>%
  kable(format = "markdown", digits = 3)

# Pairwise differences between political groups, within each contrast
contrast(coefs_pol, "revpairwise", by = "contrast") %>%
  kable(format = "markdown", digits = 3)
```

> Enter your interpretation here: Democrats vote more than rebublicans, republicants more than independents, and independents more than others. 

### Question 6.2: Education level differences {.unnumbered}

Adapt the code from 6.1 to examine differences across **education levels** instead of political groups. Run it and interpret.

::: {.callout-tip collapse="true" title="Hint"}
Replace `pol_ident` with `educ` in the `emmeans()` and `contrast()` calls.
:::

```{r}
# Estimated marginal means: political group by voter category
multi_an_pol <- emmeans(final_model, ~ educ | voter_category)

# Contrasts on the log scale (log-odds differences between voter categories, within each political group)
coefs_pol <- contrast(regrid(multi_an_pol, "log"), "trt.vs.ctrl1", by = "educ")

# Display results, organized by contrast
update(coefs_pol, by = "contrast") %>%
  kable(format = "markdown", digits = 3)

# Pairwise differences between political groups, within each contrast
contrast(coefs_pol, "revpairwise", by = "contrast") %>%
  kable(format = "markdown", digits = 3)
```

> Enter your interpretation here: These numbers support the findings earlier, more education more voting. 

------------------------------------------------------------------------

# Part 7: Write-Up

### Question 7.1 {.unnumbered}

Write a **1-paragraph results section** for your final model. Your paragraph should include:

The model we usei 
-   The model type and baseline category
-   Sample size
-   At least one key coefficient reported as an odds ratio with a confidence interval
-   The result of the likelihood ratio test for at least one predictor
-   At least one finding translated into predicted probability terms (from your plots)


::: {.callout-warning collapse="true" title="Example write-up (from a different dataset)"}
*We used a multinomial logistic regression model (baseline category: rarely/never) to predict voting frequency from age, race, gender, income, education, and political identification (N = \[n\]). The likelihood ratio test indicated that political identification was a significant predictor of voter category, χ²(\[df\]) = \[value\], p \< .001. Among \[comparison\], each additional year of age was associated with \[an increase/decrease\] in the odds of being \[category\] vs. rarely/never voting (OR = \[value\], 95% CI \[\[lower\], \[upper\]\]). Predicted probability plots showed that \[describe key pattern from your visualizations\].*
:::

> We used a multinomial logistic regression model (baseline category: rarely/never) to predict voting frequency from age, race, gender, income, education, and political identification (N= 8,327). We find that our best model incorporates all of the above variables and that age, education, and political identification has a significant effect on how likely people are to vote. For example for each increase in age by a rear, respondents are 6% more likely to be voting (odds ration 1: 1.06). The predicted probabiluty someone who went to college always votes is 30% compared to less than 20% with high school or less education. According to the Type III likelihood ratio test, age was the most significant predictor of voter category (χ²(2) = 638.30, p < 0.001).





